Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-09-28 review/response.Rmd

### Major comment

-   "Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign...However, this creates difficulties for comparing the optimisers since the end bases will be symmetric to the origin. A sign flipping step is conducted..." and the "Reconciling the orientation" section

    -   For 1D projections, flipping the sign is sufficient. For p-D projections of $\mathbb{R}^n$​, one may need a little more. To my knowledge most index functions are invariant to any rotations that happens within the p-subspace. This makes me wonder that in general, one may want to align two bases so that their difference does not contain any "within-subspace" components. In other words, one may want to compare representations in the Grassmannian manifold (space of k-subspaces) instead of in the Stiefel manifold (space of orthonormal k-frames).

\textcolor{red}{Your comment involves two different, and quite separate aspects, of projection pursuit. One is that index functions should be rotation invariant, that within the same plane any rotation of the index should have same value. There are exceptions where rotation variant indexes can still be very useful, see discussion in Laa and Cook (2020). And yes, operating on a Stiefel manifold can help - actually this is a work in progress to add to the tourr package. The same optimisation diagnostics reported in this paper should apply to Stiefel manifolds in addition to Grassmann manifolds.}

\textcolor{red}{The second element of your comment refers to the direction of projection vectors. The index value will be the same regardless of direction, but the representation of a vector as a dot on the surface of a sphere means that it may be located on either side depending on the direction. Technically it is simply that the signs of the coefficient are opposite. You encounter this in many different analyses, including PCA. In PCA, there is a convention applied to force the first element of the project vector to be positive. That is analogous to what we are doing with the flipping operation. The text in section "Reconciling the orientation" has been updated to reflect this explanation.}

@@

### Minor comment

-   The software seems to be specific to projection pursuit. I wish the authors could comment more on its potential generality to other optimizations. For example, can one generalizes the tour method to visualize any parameter trace of an optimizer?

\textcolor{red}{It is feasible to generalise the tour method for diagnosing other optimisers, given that the parameter space can be simulated. In our examples, the orthonormality constraint regulates the parameter space of the projection basis ($p \times d$) to be a sphere, or torus, which can be simulated by the geozoo package. In other problems the parameter space may take different forms in the $p$-dimensional space. Nevertheless, this can be visualised and the ferrn package should be possible to use. Researchers need first to create a data collection object in a tibble structure, analgous to the one described in our paper, that tracks the parameter during the optimisation.  Then the ferrn diagnostic plots could be used. Some additional examples of the data collection objects that we created are included in the ferrn package and should be referred to as the format for inputs into the explore functions. We have added several sentences in the conclusion explaining this.}

@@

-   Figure 4 "...All the bases in PD have been flipped for easier comparison of the final bases..."

    -   This is related to the major comment. In the Figure, PD flipped in sign but stopped at a basis representation similar to the destination of the CRS. Flipping the sign back, does it mean that PD and CRS diverged but stopped at the similar subspace with opposite basis representations? Also, given that PD and CRS have the same starting bases and one is flipped, why wouldn't the dashed line pass the origin (gray dot) in the figure?

\textcolor{red}{Thanks for sharing your interpretation on the effects of flipping, that's right! The gray dot should land on the dashed line and this was a mistake in the code. This has now been corrected in the package and Figure 4 has been updated accordingly.}

@@

### Writing

-   "Along with the bases recorded during the optimisation and a zero basis, the parameters (centre and radius) of the 2D space can be estimated by PCA. The centre of the circle is the first two PC coordinates of the zero matrix, and the radius is estimated as the largest distance from the centre to the basis."

    -   It took me multiple passes to get what this means. Perhaps instead starting with something like: "We use PCA to project and visualize the parameters/bases in 2D. The centre of the 2D view is..."?

\textcolor{red}{The sentence has been re-phrased as suggested.}

@@

1.  Should refer to earlier work on scagnostic indices, page 3

\textcolor{red}{Two references of Wilkinson work on scagnostics have been added. }

2.  figure 2 caption "using the holes index" "and a boxplot". What is the colour scale for the points? The text related to figure 2 in the bottom of page 5 should mention the dataset and the index, giving references if needed.

\textcolor{red}{The caption has now been updated and explanation for the color scale is added."}

@@

4.  In Figure 3 would be better if the y-axes scales were the same for the two graphs.

\textcolor{red}{Fixed.}

5.  What is the star in figures 4 and 5? This is not mentioned in the text or caption

\textcolor{red}{The black star represents the theoretical best basis the optimisers aiming to find. This has been added in the caption of Figure 4, but not Figure 5 as this refers back to Figure 4 in the caption.}

@@

6.  In the text you say things like "details=TRUE" and "animate=TRUE" but what functions are you referring to?

\textcolor{red}{"details=TRUE" and "animate=TRUE" refers to the function "explore space pca". These two arguments are now mentioned in the implementation section under function "explore space pca", instead of subsection Diagnostic 3a and 3b.}

7.  In Figure 9, it seems CRS is doing better here than in Figure 8 -- maybe this could be mentioned.

@@

8.  "not the case for all the indexes" missing a "." page 11

\textcolor{red}{Added.}

9.  For the Kolmogorov index, why are you using randomly generated data instead of normal quantiles?

\textcolor{red}{The Kolmogorov Smirnoff index here looks at the difference between a theoretical pdf and an empirical one based on the data. In our case we are not assuming normality on the theoretical pdf and that is the reason why we are not using normal quantiles. To make it clearer, we change the wording from "Normal Kolmogorov Index" to "Kolmogorov Index".}

### 2021-09-28 paper/zhang-cook-laa-langrene-menendez.Rmd
add css to fix bullet point line spacing

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics 
    url: Melbourne, Australia
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    url: Vienna, Austria
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    url: Melbourne, Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  patricia.menendez@monash.edu 
abstract: >
  A guided tour helps to visualise high-dimensional data by showing low-dimensional projections along a projection pursuit optimisation path. Projection pursuit is a generalisation of principal component analysis, in the sense that different indexes are used to define the interestingness of the projected data. While much work has been done in developing new indexes in the literature, less has been done on understanding the optimisation. Index functions can be noisy, might have multiple local maxima as well as an optimal maximum, and are constrained to generate orthonormal projection frames, which complicates the optimization. In addition, projection pursuit is primarily used for exploratory data analysis, and finding the local maxima is also useful. The guided tour is especially useful for exploration, because it conducts geodesic interpolation connecting steps in the optimisation and shows how the projected data changes as a maxima is approached. This work provides new visual diagnostics for examining a choice of optimisation procedure, based on the provision of a new data object which collects information throughout the optimisation. It has helped to diagnose and fix several problems with projection pursuit guided tour. This work might be useful more broadly for diagnosing optimisers, and comparing their performance. The diagnostics are implemented in the R package [ferrn](https://github.com/huizezhang-sherry/ferrn).
keywords: graphics, multivariate, optimization
preamble: >
  \usepackage{mathtools}
  \usepackage[ruled, linesnumbered]{algorithm2e}
type: package
csl: "biostats.csl"
creative_commons: CC BY
output: 
  rjtools::rjournal_web_article
bibliography: zhang-cook-laa-langrene-menendez.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      root.dir = here::here(), 
                      fig.path = here::here("figs/"), 
                      cache = TRUE,
                      fig.align = "center",
                      fig.height = 3,
                      fig.retina = 5,
                      out.width = "100%",
                      fig.cap = TRUE)
```

>>>>

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics 
    url: Melbourne, Australia
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    url: Vienna, Austria
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    url: Melbourne, Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  patricia.menendez@monash.edu 
abstract: >
  A guided tour helps to visualise high-dimensional data by showing low-dimensional projections along a projection pursuit optimisation path. Projection pursuit is a generalisation of principal component analysis, in the sense that different indexes are used to define the interestingness of the projected data. While much work has been done in developing new indexes in the literature, less has been done on understanding the optimisation. Index functions can be noisy, might have multiple local maxima as well as an optimal maximum, and are constrained to generate orthonormal projection frames, which complicates the optimization. In addition, projection pursuit is primarily used for exploratory data analysis, and finding the local maxima is also useful. The guided tour is especially useful for exploration, because it conducts geodesic interpolation connecting steps in the optimisation and shows how the projected data changes as a maxima is approached. This work provides new visual diagnostics for examining a choice of optimisation procedure, based on the provision of a new data object which collects information throughout the optimisation. It has helped to diagnose and fix several problems with projection pursuit guided tour. This work might be useful more broadly for diagnosing optimisers, and comparing their performance. The diagnostics are implemented in the R package [ferrn](https://github.com/huizezhang-sherry/ferrn).
keywords: graphics, multivariate, optimization
preamble: >
  \usepackage{mathtools, longtable}
  \usepackage[ruled, linesnumbered]{algorithm2e}
type: package
csl: "biostats.csl"
creative_commons: CC BY
output: 
  rjtools::rjournal_web_article:
    css: "style.css"
bibliography: zhang-cook-laa-langrene-menendez.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      root.dir = here::here(), 
                      fig.path = here::here("figs/"), 
                      cache = TRUE,
                      fig.align = "center",
                      fig.height = 3,
                      fig.retina = 5,
                      out.width = "100%")
```

### 2021-09-28 paper/zhang-cook-laa-langrene-menendez.tex
add css to fix bullet point line spacing

Several annotations have been made to help understanding this plot. As
mentioned previously, the original basis space is a high-dimensional
sphere and random bases on the sphere can be generated via the
\CRANpkg{geozoo} \citep{geozoo} package. We use PCA to estimate the 2D
space that is best suited to visualise the bases. The centre of the 2D
view is the first two PC coordinates of the zero matrix, and the radius
is estimated as the largest distance from the centre to the basis. The
theoretical best basis is known for simulated data and can be labelled
to compare how close the end basis found by the optimisers is to the
theoretical best. Various aesthetics, i.e.~size, alpha (transparency),
and colour, are applicable to emphasize critical elements and adjust for
the presentation. For example, anchor points and search points are less
important and hence a smaller size and alpha are used. Alpha can also be
applied on the interpolation paths to show the start to finish from
transparent to opaque.

@@

This paper has provided several visual diagnostics that can be used for
understanding a complex optimisation procedure and are implemented in
the ferrn package. The methods were illustrated using the optimisers
available for projection pursuit guided tour. Here the constraint is the
orthonormality condition of the projection bases. The approach used
broadly applies to other constrained optimisers, by saving the path of
the optimiser in a form required to input into the ferrn package. These
high-dimensional data objects can then be viewed using the tour.

The progressive optimisation of a target function and it's coverage of
the search space can be viewed in both reduced 2D space and the full
space. These visualisation can lead to insights for evaluating and
comparing the performance of multiple optimisers in a specific problem.
They can provide better understanding of existing methods or motivate
the development of new approaches. For example, we have compared how
three optimisers perform when maximising a non-smooth index function and
illustrated how the pseudo-derivative search fails in this setting. The
observations from our experiments have also been translated into
improved optimisation methods for the guided tour, e.g.~we introduced
the option to interrupt the search if a better basis is found along the
path.

>>>>

Several annotations have been made to help understanding this plot. As
mentioned previously, the original basis space is a high-dimensional
sphere and random bases on the sphere can be generated via the
\CRANpkg{geozoo} \citep{geozoo} package. We use PCA to project and
visualize the parameters/bases in 2D. The centre of the 2D view is the
first two PCs of the data matrix. It theoretically should be a circle,
but may have some irregular edges due to finite sampling. Thus the edge
is smoothed by using a radius estimated as the largest distance from the
centre to any basis. In the simulation, the theoretical best basis is
known and can be labelled to compare how close to this that the
optimisers stopped. Various aesthetics, i.e.~size, alpha (transparency),
and colour, are applicable to emphasize critical elements and adjust for
the presentation. For example, anchor points and search points are less
important and hence a smaller size and alpha are used. Alpha can also be
applied on the interpolation paths to show the start to finish from
transparent to opaque.

@@

This paper has provided several visual diagnostics that can be used for
understanding a complex optimisation procedure and are implemented in
the ferrn package. The methods were illustrated using the optimisers
available for projection pursuit guided tour. Here the constraint is the
orthonormality condition of the projection bases, which corresponds to
optimisation over spheres and torii. The approach described broadly
applies to other constrained optimisers. Although the manifold in
\(p\)-space might be different the diagnostic techniques are the same. A
researcher would begin by saving the path of the optimiser in a form
required to input into the ferrn package, as described in this paper.
One might generally make more samples from the constrained space upon
which to assess and compare the optimisation paths. These
high-dimensional data objects can then be viewed using the tour.

The progressive optimisation of a target function and it's coverage of
the search space can be viewed in both reduced 2D space and the full
space. These visualisations can lead to insights for evaluating and
comparing the performance of multiple optimisers operating on the same
task. They can provide better understanding of existing methods or
motivate the development of new approaches. For example, we have
compared how three optimisers perform when maximising a non-smooth index
function and illustrated how the pseudo-derivative search fails in this
setting. The observations from our experiments have also been translated
into improved optimisation methods for the guided tour, e.g.~we
introduced the option to interrupt the search if a better basis is found
along the path.
